---
title: "Racial Wage Gap"
author: "Quincy Troutman"
date: "CIS365 4/6/2021"
output:
  html_document: default
subtitle: Developing Analytics Applications in R
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract
The racial wage gap (RWG), or in other words a comparison of pay between white 
people and people of color who have the same job and qualifications (Scale, 2020),
in the technology sector has generated many concerns. PayScale.com surveyed 
52,000 workers in the tech industry and found the African American stem workers 
made almost $0.13 on the dollar less than white workers, while Asian workers make 
$0.23 on the dollar more than white workers (Perez, 2019). Many variables are 
said to cause this wage gap, such as cognitive skills and raw intelligence. The 
problem is that do these variables actually explain a wage gap or is it over dramatized. 

RWG in the tech industry is important in many ways. Racism is illegal in the 
workplace but, even more important is that effected citizens could lose thousands
of dollars in wages. According to PayScale’s survey, Black men make about $73,000
a year and white men make around $74,500 a year. An extra $1,500 a year could 
help an employee pay bills or go on a vacation and help improve that employees 
motivation and performance within the company. 

The research question that I developed was 

* “Is there a racial wage gap in the technology sector?”. 

I will be getting my data from the developer survey results from 2019 that is located
on Stack Overflows website will be the main source of data for this project. 
This data shows survey results of 90,000 different developers. The 20-minute 
survey shows data from computer language preferences to life outside of work. 
Even though the data is overly broad,I will be able to narrow down the data to 
help show if there is a racial wage gap in the technology sector and answer my 
research question.

In order to Analyze this data I used the Tukey HSD Model.Tukey HSD (honest significant difference) test is a statistical tool used to determine if the relationship between two sets of data is statistically significant. (Beck, 2020) 

Using this model I found that the relationship between Ethnicity and Pay is not statistically significant. This means that it is likely to be explained solely by chance or random factors.

From a test I would recommend using a completely different model such as Ancova along with Anova. 
From using Tukey, I found that the relationship between Ethnicity and Average pay in the technology sector is not significant according to my data.

# Data

As said before, I sourced my data from the developer survey results from 2019.this is located on Stack Overflows website at https://insights.stackoverflow.com/survey/2019.

## Devloping my data

First I uploaded the developer survey results to the R File and saved it in "All Data"

```{r My working Directory, echo=TRUE, cache=TRUE,  results='hide'}
mywd <- "C:/Users/Quinc/OneDrive/Murray State/Spring 2021/CIS 361"
setwd(mywd)


myfile <- "survey_results_public.csv"

Alldata <- read.csv(file = myfile, header = TRUE, sep = ",")

```

These are the packages that I used throughout my research. 

```{r packages, echo=TRUE, cache=TRUE,  results='hide'}
library(tidyverse)
library(ggpubr)
library(rstatix)
library(knitr)
library(stargazer)
library(ggplot2)
library(vctrs)
library(plyr)
library(dplyr)
library(rmarkdown)

```
 
 
Next I took 'All Data' and picked what data I would use.
I chose to use the columns 'Respondent', 'Employment', 'Country', 'Ethnicity', 'DevType', 
'YearsCode', 'CompTotal', 'WorkWeekHrs', 'Age', 'Gender'. I stored these values in 'SmallData'.


```{r SmallData, echo=FALSE, cache=TRUE, tidy=TRUE,  results='hold'}
#Pulling the Columns I want to see.
SmallData <- dplyr::select(Alldata, Respondent, Employment, Country, Ethnicity, DevType
                    , YearsCode, CompTotal, WorkWeekHrs, Age, Gender)
SmallData <-na.omit(SmallData)

knitr::kable(SmallData[1:10,1:10], format="html", round(digits = 2), caption = "'SmallData' Data Frame (First ten Observations).")

```



After that, I filtered the 'Country' Column to only show respondents from the United States.
Also, I narrowed 'DevType' to the respondents first choice in the list of items Provided.

I Stored this Information in the 'FromUSA' Data Frame as shown below.



```{r FromUSA, echo=FALSE, cache=TRUE, tidy=TRUE,  results='hold'}
#Filtering columns to get the right Data.
FromUSA <- SmallData %>% filter(Country == 'United States', Employment == 'Employed full-time')

#Remove rows with null values
FromUSA <- na.omit(FromUSA)

#Narrowing the scope of DevTypes
FromUSA[] <- FromUSA %>% 
  group_by(DevType) %>%
  arrange(DevType)
i <- 0
while (i <= 533){
  FromUSA$DevType[i] <- "Academic researcher"
  i = i + 1
}
while (i <= 1490){
  FromUSA$DevType[i] <- "Data or business analyst"
  i = i + 1
}
while (i <= 2000){
  FromUSA$DevType[i] <- "Data scientist or machine learning specialist"
  i = i + 1
}
while (i <= 2943){
  FromUSA$DevType[i] <- "Database administrator"
  i = i + 1
}
while (i <= 3620){
  FromUSA$DevType[i] <- "Designer"
  i = i + 1
}
while (i <= 12083 ){
  FromUSA$DevType[i] <- "Developer"
  i = i + 1
} 
while (i <= 12200 ){
  FromUSA$DevType[i] <- "DevOps specialist"
  i = i + 1
} 
while (i <= 12221 ){
  FromUSA$DevType[i] <- "Educator"
  i = i + 1
}
while (i <= 12433  ){
  FromUSA$DevType[i] <- "Engineer"
  i = i + 1
}
while (i <= 12465 ){
  FromUSA$DevType[i] <- "Product manager"
  i = i + 1
}
while (i <= 12488  ){
  FromUSA$DevType[i] <- "Scientist"
  i = i + 1
}
while (i <= 12537 ){
  FromUSA$DevType[i] <- "Senior executive/VP"
  i = i + 1
}
while (i <= 12541 ){
  FromUSA$DevType[i] <- "Student"
  i = i + 1
} 

FromUSA$YearsCode[grepl("More than 50 years",FromUSA$YearsCode)]<-'55'
FromUSA$YearsCode[grepl("Less than 1 year",FromUSA$YearsCode)]<-'1'
FromUSA$YearsCode <- as.numeric(FromUSA$YearsCode)
##head(FromUSA)
knitr::kable(FromUSA[1:10,1:10], format="html", round(digits = 2), caption = "Respondents From The United States of America (First ten Observations)")


```

Next I wanted to see what the data looked like when it was grouped by 'Devtype' then Ethnicity, so I created a Data Frame to show that and called it 'DevTypeVsEthnicity'.

This shows the number of people per each of the 9 'Ethnicity' in one of the 14 selected 'DevTypes', along with the average of pay, hours per work week, Age and years experience.


```{r DevTypeVsEthnicity, echo=FALSE, cache=TRUE, tidy=TRUE,  results='hold'}
DevTypeVsEthnicity <- FromUSA %>% 
  select(c(DevType, Ethnicity, WorkWeekHrs, CompTotal, Age, YearsCode)) %>%
  group_by(DevType, Ethnicity) %>%
  dplyr::summarise(NumberOfPeople = n(), AveragePay = mean(CompTotal), AverageHrsPerWeek = mean(WorkWeekHrs), AverageAge = mean(Age), AverageYearsExperience = mean(YearsCode))
#head(DevTypeVsEthnicity)
knitr::kable(DevTypeVsEthnicity[1:10,1:7], format="html", round(digits = 2),  caption = "Averages of Hours Per Week, Average Age, and Average Years Experience Grouped by Dev Type and Ethnicity (First ten Observations)")
```

## Regression

After that was done, I moved on to my analysis. I started developing an Anova test using the tukey model. In order to do this I needed to get summaries of important data that I need for this process. So, I created these tables using 'DevTypeVsEthnicity' Data Frame.

### Summary Tables

```{r Regression, echo=FALSE, cache=TRUE, tidy=TRUE,  results='hold'}

#DF (Ethnicity, DevType, AveragePay)
DevType_Ethnicity_Pay<- DevTypeVsEthnicity %>%
  dplyr::select(Ethnicity, DevType, AveragePay)

knitr::kable(DevType_Ethnicity_Pay[1:10,1:3], format="html", round(digits = 2),  caption = "Averages of Pay Grouped by Ethnicity and Devtype (First ten Observations)")

#DF (Ethnicity, DevType,AverageYearsExperience)
DevType_Ethnicity_Exp <- DevTypeVsEthnicity %>%
 dplyr::select(Ethnicity, DevType,AverageYearsExperience)

knitr::kable(DevType_Ethnicity_Exp[1:10,1:3], format="html", round(digits = 2),  caption = "Average Year Experience Grouped by Ethnicity and Devtype (First ten Observations)")

#Summary of mean pay Group by DevType, Ethnicity
AnovaDevtypeEthPay <- DevType_Ethnicity_Pay %>% 
  group_by(DevType, Ethnicity) %>%
  get_summary_stats(AveragePay, type = "mean")

knitr::kable(AnovaDevtypeEthPay[1:10,1:5], format="html", round(digits = 2),  caption = "Summary of Pay Grouped by Ethnicity and Devtype (First ten Observations)")

#Summary of mean Experience Group by DevType, Ethnicity
AnovaDevtypeEthExp <- DevType_Ethnicity_Exp %>% 
 #facet_wrap(~DevType) %>%
  group_by(DevType, Ethnicity) %>% 
   get_summary_stats(AverageYearsExperience, type = "mean")

knitr::kable(AnovaDevtypeEthExp[1:10,1:5], format = "html", round(digits = 2),  caption = "Summary of Year Experience Grouped by Ethnicity and Devtype (First ten Observations)")

```

Then, I created a Bar chart showing my data. I did this to hopefully get a better understanding of what I was working with and understand the differences among my date. 

### Summary Bar Graph

```{r Column Charts,  fig.width = 15, fig.height = 10, echo=FALSE, cache=TRUE, tidy=TRUE,  results='hold'}

options(scipen=999)

AnovaDevtypeEthPayBar <-ggplot(DevType_Ethnicity_Pay, aes(x = Ethnicity,                                                           y = AveragePay, fill = Ethnicity)) + 
  geom_col() + facet_wrap(~DevType)+ labs(title="Average Pay Per DevType", subtitle="Average Annual Pay in USD",
       y="Average Pay", x="Ethnicity", caption="Figure 1. Average Pay") +
  scale_x_discrete(labels = c('A','B','C','D','E','F','G','H','I'))

AnovaDevtypeEthEXPBar <-ggplot(DevType_Ethnicity_Exp, aes(x = Ethnicity,                                                           y = AverageYearsExperience, fill = Ethnicity, label = Ethnicity)) + 
  geom_col() + facet_wrap(~DevType)+ labs(title="Average Experience Per DevType", subtitle="Average Experience in Years",
       y="Average Experience", x="Ethnicity", caption="Figure 1. Average Experience") +
  scale_x_discrete(labels = c('A','B','C','D','E','F','G','H','I'))
  
plot(AnovaDevtypeEthPayBar)
plot(AnovaDevtypeEthEXPBar)



```
  

Now that I have a good understanding of my data I can begin with my Anova Analysis 

# Methodology 

My analysis of my data was done using the Tukey_HSD test. This test is a statistical tool
used to determine if the relationship between two sets of data is statistically significant.  

## Tukey Analysis

I made four models. The main model of interest would be the 'Pay and Ethnicity Model'. 

### Pay And DevType

```{r Anova, echo=FALSE,fig.width = 9, fig.height = 10, cache=TRUE, tidy=TRUE,  results='hold'}

set.seed(1234)

#p-Value = 0.273 (Pay to DevType)
model <- aov(mean~DevType, data=AnovaDevtypeEthPay)

summary(model)
#turkey for model (Pay to DevType)
Turkey<-TukeyHSD(model, conf.level=.95, )
#Turkey plot (Pay to DevType)
knitr::kable(Turkey[], format = "html", round(digits = 2),  caption = "Pay to DevType")
plot(Turkey, las=2) 

```

### Experience and DevType

```{r Anova2, echo=FALSE,fig.width = 9, fig.height = 10, cache=TRUE, tidy=TRUE,  results='hold'}
set.seed(1234)
#p-Value = .00864 (Experience to DevType)
model2 <- aov(mean~DevType, data=AnovaDevtypeEthExp)

summary(model2)
#turkey for model 2(Experience to DevType)
Turkey2<-TukeyHSD(model2, conf.level=.95)
#Turkey plot (Experience to DevType)
knitr::kable(Turkey2[], format = "html", round(digits = 2),  caption = "Pay to DevType")
plot(Turkey2, las=2)

```

### Pay and Ethnicity

```{r Anova3, echo=FALSE, fig.width = 9, fig.height = 10,cache=TRUE, tidy=TRUE,  results='hold'}

#p-Value = .28 (Pay to Ethnicity)
model3 <- aov(mean~Ethnicity, data=AnovaDevtypeEthPay)

summary(model3)
#turkey for model 3(Pay to Ethnicity)
Turkey3<-TukeyHSD(model3, conf.level=.95)
#Turkey plot (Pay to Ethnicity)
knitr::kable(Turkey3[], format = "html", round(digits = 2),  caption = "Pay to DevType")
plot(Turkey3, las=2)
```
## Experience and Ethnicity

```{r Anova4, echo=FALSE, fig.width = 9, fig.height = 10, cache=TRUE, tidy=TRUE,  results='hold'}

#p-Value = .339 (Experience to Ethnicity)
model4 <- aov(mean~Ethnicity, data=AnovaDevtypeEthExp)

summary(model4)
#turkey for model 4(Experience to Ethnicity)
Turkey4<-TukeyHSD(model4, conf.level=.95)
#Turkey plot (Experience to Ethnicity)
knitr::kable(Turkey4[], format = "html", round(digits = 2),  caption = "Pay to DevType")
plot(Turkey4,las=2)
```



# Results

Pay And DevType.
* P-value = 0.273
* Not Significant

Experience and DevType.
* P-value = 0.28
* Not Significant


Pay and Ethnicity.
* P-value = 0.339
* Not Significant

Experience and Ethnicity. 
* P-value = .00893
* Significant


# Conclusion
According to my analysis, my data showed that the technology sector in terms of a Pay gap can be explained solely by chance or random factors. This means that my data shows that equal pay for equal work is present.


# References

Peffers, K., Tuunanen, T., Rothenberger, M. A., & Chatterjee, S. (2007). A design science research methodology for information systems research. Journal of management information systems, 24(3), 45-77.

2020 Racial Wage Gap - Compensation Research from PayScale. PayScale. (2021). Retrieved 24 February 2021, from https://www.payscale.com/data/racial-wage-gap#:~:text=The%20controlled%20racial%20wage%20gap%20(RWG)%20is%20a,is%20alive%20and%20well%20in%20the%20U.S.%20workforce.

Perez, T. (2019, January 7). The state of wage inequality in tech in 2018. PayScale. https://www.payscale.com/data/racial-wage-gap-in-tech

Petre, M. (2019). Contributions of Skills to the Racial Wage Gap. Journal of Human Capital, 13(3), 479–518. https://doi-org.ezproxy.waterfield.murraystate.edu/10.1086/704322

Beck, K. (2020, December 21). What is the tukey hsd test? Retrieved May 02, 2021, from https://sciencing.com/what-is-the-tukey-hsd-test-12751748.html










